### 2014-05-09 writeups/context-dependence.tex
Swapping h for ell. Closes #1.

The common framework of the above examples is a one-dimensional grid of $n$ sites,
and each site is labeled with one of a finite set of states.
These change with time -- so, write $X_i(t)$ for the state of the $i^\mathrm{th}$ site at time $t$,
where each $X_i(t) \in \calS$, the set of possible states.
(To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
We define a \textit{pattern} to mean a contiguous sequence of states, and we let $|u|$ denote the length of the pattern $u$.
The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
The sequence-level transition process is the independent superposition of all of the individual pattern mutation processes.
Thus given two patterns $u$ and $v$ of common length $\ell$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+\ell}(t)) = (u_1, \ldots, u_\ell)$ then
\[
\P\{(X_i(t+dt), \ldots, X_{i+\ell}(t+dt)) = (v_1, \ldots, v_\ell) = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
We call such a $(\mu,u,v)$ a ``transition triple''.
The set of all transition triples $\calT = \{ (\mu^j,u^j,v^j) \}_{j=1}^{n_T}$
determines the time evolution of the process.

@@

\paragraph{The generator matrix}
For clarity, we should define how the set of transition rates determines the transition rate matrix,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x$ jumps to state $y$.
% Let $x_i^{(\ell)} = (x_i, x_{i+1}, \ldots, x_{i+\ell-1})$ be the subsequence of length $\ell$ beginning at location $i$, and
For each $1\le i \le n$, pattern length $\ell \ge 0$, and patterns $u,v \in \calS^\ell$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < \ell \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < \ell, and \\
  x_k = y_k \quad &\text{for } k\ge i+\ell ,
\end{cases}
\]
i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+\ell-1$, where $x$ matches with $u$ and $y$ matches with $v$.
(Interpet these indices modulo $n$, even for short sequences, a point we will come back to.)
We'll also need the set of defined transitions in $\calT$
that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
  G(n)_{x,y} = \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This is just to say that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
that could change $x$ into $y$.
In principle, this gives us the transition probabilities for the process:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
\end{align}
but in practice the size of these matrices can make a direct application of this formula inefficient.
This paper proposes an alternate strategy.

@@

First, it will be useful to have a brief description of how to simulate such a process.
Briefly, we can put down a Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$,
and then resolve each possible change in temporal order.
\todo{OK, rad.}
We situate ``changes'' at the beginning of the matching pattern,
so $\mu_*$ can be defined as the maximum sum of rates matching  the start of a sequence:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(1,x,y)} \mu_j \st x \in \calS^n \right\} .
\]
\todo[inline]{OK, I'm getting confused now. From the above, I would have thought that $\mu_*$ was the overall rate of mutation, but now it's only happening from pattern x?}
Now sample possible changes: let $c_{ij}$ denote the $j^\text{th}$ possible change at site $i$,
so that
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a rate $\mu_*$ Poisson process.
\todo[inline]{Further confusion: now $\mu_*$ is just the rate at a single site?}
Then, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $s_{ij} = \min \{ s_{k\ell} : s_{k\ell}>s \}$.
The process then jumps to a new state at times $s_{ij}$ with probability proportional to the corresponding rate.
In other words, let
\[
q_k = \begin{cases}
  \mu_k/\mu_* \qquad & \text{if } x_i^{(|u_k|)} = u_k  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
and let $q_0 = 1-\sum_{k=1}^{n_T} q_k$.
By construction of $\mu_*$, $q_0\ge 0$.)
Now choose transition $k$ with probability $q_k$, and if $k>0$, replace $x_i^{|u_k|}$ with $v_k$.

>>>>

The common framework of the above examples is a one-dimensional grid of $n$ sites,
and each site is labeled with one of a finite set of states.
These change with time -- so, write $X_i(t)$ for the state of the $i^\mathrm{th}$ site at time $t$,
where each $X_i(t) \in \calS$, the set of possible states.
(To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
We define a \textit{pattern} to mean a contiguous sequence of states, and we let $|u|$ denote the length of the pattern $u$.
The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
The sequence-level transition process is the independent superposition of all of the individual pattern mutation processes.
Thus given two patterns $u$ and $v$ of common length $h$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+h}(t)) = (u_1, \ldots, u_h)$ then
\[
\P\{(X_i(t+dt), \ldots, X_{i+h}(t+dt)) = (v_1, \ldots, v_h) = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
We call such a $(\mu,u,v)$ a ``transition triple''.
The set of all transition triples $\calT = \{ (\mu^j,u^j,v^j) \}_{j=1}^{n_T}$
determines the time evolution of the process.

@@

\paragraph{The generator matrix}
For clarity, we should define how the set of transition rates determines the transition rate matrix,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x$ jumps to state $y$.
% Let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ be the subsequence of length $h$ beginning at location $i$, and
For each $1\le i \le n$, pattern length $h \ge 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < h, and \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+h-1$, where $x$ matches with $u$ and $y$ matches with $v$.
(Interpet these indices modulo $n$, even for short sequences, a point we will come back to.)
We'll also need the set of defined transitions in $\calT$
that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
  G(n)_{x,y} = \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This is just to say that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
that could change $x$ into $y$.
In principle, this gives us the transition probabilities for the process:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
\end{align}
but in practice the size of these matrices can make a direct application of this formula inefficient.
This paper proposes an alternate strategy.

@@

First, it will be useful to have a brief description of how to simulate such a process.
Briefly, we can put down a Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$,
and then resolve each possible change in temporal order.
\todo{OK, rad.}
We situate ``changes'' at the beginning of the matching pattern,
so $\mu_*$ can be defined as the maximum sum of rates matching  the start of a sequence:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(1,x,y)} \mu_j \st x \in \calS^n \right\} .
\]
\todo[inline]{OK, I'm getting confused now. From the above, I would have thought that $\mu_*$ was the overall rate of mutation, but now it's only happening from pattern x?}
Now sample possible changes: let $c_{ij}$ denote the $j^\text{th}$ possible change at site $i$,
so that
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a rate $\mu_*$ Poisson process.
\todo[inline]{Further confusion: now $\mu_*$ is just the rate at a single site?}
Then, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $s_{ij} = \min \{ s_{kh} : s_{kh}>s \}$.
The process then jumps to a new state at times $s_{ij}$ with probability proportional to the corresponding rate.
In other words, let
\[
q_k = \begin{cases}
  \mu_k/\mu_* \qquad & \text{if } x_i^{(|u_k|)} = u_k  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
and let $q_0 = 1-\sum_{k=1}^{n_T} q_k$.
By construction of $\mu_*$, $q_0\ge 0$.)
Now choose transition $k$ with probability $q_k$, and if $k>0$, replace $x_i^{|u_k|}$ with $v_k$.
